---
title: "Final Project:  Clustering Analysis"
author: "Katherine M. Prioli"
date: "December 05, 2019"
output:
  pdf_document: default
  html_document: default
geometry: margin = 0.5in
---


For the unsupervised component of this analysis, data was subset to variables with $\ge 90 \%$ non-missing values, then all cases with missing values were dropped.  Since this approach involves visual analysis via dendrograms, the data was subset via a random 5% sample within each BMI category to allow for sensible, interpretable dendrograms.  The analytic dataset thus comprised `r dim(nhanes_unsup_mat)[1]` cases having `r dim(nhanes_unsup_mat)[2]` variables.  To generate dendrograms for agglomerative clustering (AGNES) and divisive clustering (DIANA), a dissimilarity matrix was calculated based on Gower distance.  The Gower distance was chosen because the variables in this dataset are of mixed type (i.e., some continuous, some ordinal, and some nominal).

Descriptives for the dissimilarity matrix:

```{r dissim, echo = FALSE, paged.print = FALSE, fig.width = 10, fig.height = 6}
dissim_mat <- vegdist(nhanes_unsup_mat, method = "gower")
summary(dissim_mat) %>% 
  round(digits = 2) %>% 
  tidy() %>% 
  setNames(., c("Min", "Q1", "Median", "Mean", "Q3", "Max")) %>% 
  kable(format = "markdown")
```

```{r agnes_diana, echo = FALSE, paged.print = FALSE, fig.width = 10, fig.height = 6}
nhanes_agnes <- hclust(d = dissim_mat, method = "ward.D2")   # ward.D2 performs agglomerative per hclust documentation

nhanes_diana <- diana(x = (dissim_mat %>% as.matrix()), diss = TRUE, keep.diss = TRUE) %>% 
  as.hclust()
```

Because the data is likely noisy, AGNES relied on complete (i.e., maximum distance) linkage between clusters.  The radial dendrogram for AGNES is shown below, with leaves colored to indicate BMI category as follows:

* Underweight = Yellow
* Healthy weight = Green
* Overweight = Orange
* Obese = Red

```{r agnes_dend, warning = FALSE, message = FALSE, echo = FALSE, paged.print = FALSE, fig.width = 11, fig.height = 11}
ggthemr("flat", spacing = 0.1)

unsup_labs_reord_ag <- unsup_labs[nhanes_agnes$order] %>%    # Reordering leaf labels
  str_sub(start = 1L, end = 1L) %>% 
  as_tibble() %>% 
  mutate(BMI_cat = case_when(   # Modifying leaf labels to show BMI category abbreviations
    value == "1" ~ "UW",        # Underweight
    value == "2" ~ "HW",        # Healthy weight
    value == "3" ~ "OW",        # Overweight
    value == "4" ~ "OB"         # Obese
  )) %>% 
  select(BMI_cat) %>% 
  as_tibble() %>% 
  mutate(x = row_number(),
         labcolor = case_when(   # Have decided not to use text labels but keeping this here in case I change my mind
           BMI_cat == "UW" ~ "#B0B000",
           BMI_cat == "HW" ~ "#008318",
           BMI_cat == "OW" ~ "#FF8D03",
           BMI_cat == "OB" ~ "#FF0000"),
         leafcolor = case_when(
           BMI_cat == "UW" ~ "#FFFF00",
           BMI_cat == "HW" ~ "#008318",
           BMI_cat == "OW" ~ "#FF8D03",
           BMI_cat == "OB" ~ "#FF0000")) %>% 
  select(x, BMI_cat, labcolor, leafcolor)

ag_labcolors <- unsup_labs_reord_ag

agnes_dend <- nhanes_agnes %>% 
  as.dendrogram() %>% 
  set("leaves_pch", 19, col = ag_labcolors$leafcolor) %>% 
  set("leaves_col", ag_labcolors$leafcolor)

agnes_dend_ggd <- agnes_dend %>% 
  as.ggdend()

agnes_dend_gg <- ggplot(agnes_dend_ggd, labels = FALSE) +
  scale_y_reverse() +
  coord_polar(theta = "x") +
  theme(text = element_text(face = "bold"),
        axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(angle = 0)) +
  xlab("") +
  ylab("") #+
  #ggtitle("AGNES Dendrogram")
agnes_dend_gg

ggthemr("flat")
```

DIANA yielded the following dendrogram:

```{r diana_dend, echo = FALSE, warning = FALSE, paged.print = FALSE, fig.width = 11, fig.height = 11}
ggthemr("flat", spacing = 0.1)

unsup_labs_reord_di <- unsup_labs[nhanes_diana$order] %>%    # Reordering leaf labels
  str_sub(start = 1L, end = 1L) %>% 
  as_tibble() %>% 
  mutate(BMI_cat = case_when(   # Modifying leaf labels to show BMI category abbreviations
    value == "1" ~ "UW",        # Underweight
    value == "2" ~ "HW",        # Healthy weight
    value == "3" ~ "OW",        # Overweight
    value == "4" ~ "OB"         # Obese
  )) %>% 
  select(BMI_cat) %>% 
  as_tibble() %>% 
  mutate(x = row_number(),
         labcolor = case_when(   # Have decided not to use text labels but keeping this here in case I change my mind
           BMI_cat == "UW" ~ "#B0B000",
           BMI_cat == "HW" ~ "#008318",
           BMI_cat == "OW" ~ "#FF8D03",
           BMI_cat == "OB" ~ "#FF0000"),
         leafcolor = case_when(
           BMI_cat == "UW" ~ "#FFFF00",
           BMI_cat == "HW" ~ "#008318",
           BMI_cat == "OW" ~ "#FF8D03",
           BMI_cat == "OB" ~ "#FF0000")) %>% 
  select(x, BMI_cat, labcolor, leafcolor)

di_labcolors <- unsup_labs_reord_di

diana_dend <- nhanes_diana %>% 
  as.dendrogram() %>% 
  set("leaves_pch", 19, col = di_labcolors$leafcolor) %>% 
  set("leaves_col", di_labcolors$leafcolor)

diana_dend_ggd <- diana_dend %>% 
  as.ggdend()

diana_dend_gg <- ggplot(diana_dend_ggd, labels = FALSE) +
  scale_y_reverse() +
  coord_polar(theta = "x") +
  theme(text = element_text(face = "bold"),
        axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(angle = 0)) +
  xlab("") +
  ylab("") #+
  #ggtitle("DIANA Dendrogram")
diana_dend_gg

ggthemr("flat")
```

```{r clust_stats, echo = FALSE, paged.print = FALSE, fig.width = 10, fig.height = 6}
source(here("clustering_stats.R"), echo = FALSE)
```

For both AGNES and DIANA, cluster size was assessed and within-cluster sums of squares and average silhouette width were calculated based on an upper limit of 15 clusters; these latter two metrics were used to generate scree and silhouette plots.  Cluster sizes were as follows:

```{r clust_sizes, warning = FALSE, echo = FALSE, paged.print = FALSE, fig.width = 12, fig.height = 4}
maxheight = rbind(agnes_sizes$cluster_size, diana_sizes$cluster_size) %>% 
  max()
ymax = ceiling(maxheight / 5) * 5

agnes_clustsize <- ggplot(data = agnes_sizes, aes(x = cluster_no, y = cluster_size)) +
  geom_col() +
  scale_x_continuous(breaks = c(0:15)) +
  ylim(0, ymax) +
  xlab("Cluster Number") +
  ylab("Cluster Size") +
  ggtitle("AGNES")

agnes_sizedesc <- summary(agnes_sizes$cluster_size) %>% 
  round(digits = 2) %>% 
  tidy() %>% 
  setNames(., c("Min", "Q1", "Median", "Mean", "Q3", "Max")) %>% 
  t() %>% 
  tableGrob(theme = ttheme_minimal(base_size = 10, padding = unit(c(2, 2), "mm")))

diana_clustsize <- ggplot(data = diana_sizes, aes(x = cluster_no, y = cluster_size)) +
  geom_col() +
  scale_x_continuous(breaks = c(0:15)) +
  ylim(0, ymax) +
  xlab("Cluster Number") +
  ylab("Cluster Size") +
  ggtitle("DIANA")

diana_sizedesc <- summary(diana_sizes$cluster_size) %>% 
  round(digits = 2) %>% 
  tidy() %>% 
  setNames(., c("Min", "Q1", "Median", "Mean", "Q3", "Max")) %>% 
  t() %>% 
  tableGrob(theme = ttheme_minimal(base_size = 10, padding = unit(c(2, 2), "mm")))

grid.arrange(agnes_sizedesc, agnes_clustsize, textGrob(" "), diana_sizedesc, diana_clustsize, 
             nrow = 1, ncol = 5, widths = c(0.25, 0.7, 0.1, 0.25, 0.7),
             top = textGrob("Cluster Sizes, k = 15 Clusters, Hierarchical Clustering Analysis",
                            gp = gpar(fontsize = 16, fontface = "bold", col = "#34495E"),
                            x = 0.025, hjust = 0))
```

Though neither plot shows an even size distribution across clusters, AGNES' cluster sizes are less varied than those of DIANA.

```{r scree, echo = FALSE, paged.print = FALSE, fig.width = 10, fig.height = 4}
agnes_scree <- ggplot(data = agnes_statdata, aes(x = n_clusters, y = withinss)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = c(0:15)) +
  xlab("Number of Clusters") +
  ylab("Within-Cluster Sum of Squares") +
  ggtitle("AGNES")

diana_scree <- ggplot(data = diana_statdata, aes(x = n_clusters, y = withinss)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = c(0:15)) +
  xlab("Number of Clusters") +
  ylab("Within-Cluster Sum of Squares") +
  ggtitle("DIANA")

grid.arrange(agnes_scree, diana_scree, nrow = 1,
             top = textGrob("Scree Plots",
                            gp = gpar(fontsize = 16, fontface = "bold", col = "#34495E"),
                            x = 0.025, hjust = 0))
```

The AGNES scree plot shows a relatively smooth curve without strong elbows; possible but faint elbows appear at 5 and 7 clusters.  For DIANA, two obvious elbows are seen:  one at 5 clusters and one at 6.

```{r silhou, echo = FALSE, paged.print = FALSE, fig.width = 10, fig.height = 4}
agnes_sil <- ggplot(data = agnes_statdata, aes(x = n_clusters, y = asw)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = c(0:15)) +
  xlab("Number of Clusters") +
  ylab("Average Silhouette Width") +
  ggtitle("AGNES")

diana_sil <- ggplot(data = diana_statdata, aes(x = n_clusters, y = asw)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = c(0:15)) +
  xlab("Number of Clusters") +
  ylab("Average Silhouette Width") +
  ggtitle("DIANA")

grid.arrange(agnes_sil, diana_sil, nrow = 1,
             top = textGrob("Silhouette Plots",
                            gp = gpar(fontsize = 16, fontface = "bold", col = "#34495E"),
                            x = 0.025, hjust = 0))
```

Both the AGNES and DIANA silhouette plots show maximum average silhouette width at 2 clusters, which is unlikely to contain sufficient data to be meaningful.  AGNES shows a local maximum at 7 clusters, and DIANA at 6.

The AGNES scree and silhouette plots indicate that 7 clusters may be sufficient.  Six clusters were chosen for DIANA.

```{r agnes_dend_clust_init, echo = FALSE, warning = FALSE, paged.print = FALSE, fig.width = 11, fig.height = 11}
# 7 clusters for AGNES
ggthemr("flat", spacing = 0.1)

agnes_dend_clust_init <- nhanes_agnes %>% 
  as.dendrogram() %>% 
  set("leaves_pch", 19, col = ag_labcolors$leafcolor) %>% 
  set("leaves_col", ag_labcolors$leafcolor) %>% 
  set("branches_k_col", k = 7, c("#3498DB",    # Blue
                                 "#2ECC71",    # Green
                                 "#F39C12",    # Gold/orange
                                 "#E74C3C",    # Red
                                 "#D979C9",    # Pink
                                 "#9B59B6",    # Purple
                                 #"#9F83D4",    # Grape
                                 "#1ABCB4")) #,    # Teal
                                 #"#34495E"))   # Slate blue

agnes_dend_clust_init_ggd <- agnes_dend_clust_init %>% 
  as.ggdend()

agnes_dend_clust_init_gg <- ggplot(agnes_dend_clust_init_ggd, labels = FALSE) +
  scale_y_reverse() +
  coord_polar(theta = "x") +
  theme(axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  xlab("") +
  ylab("") #+
  #ggtitle("AGNES Dendrogram, k = 7 Clusters")
agnes_dend_clust_init_gg

ggthemr("flat")
```

```{r diana_dend_clust_k6, echo = FALSE, warning = FALSE, paged.print = FALSE, fig.width = 11, fig.height = 11}
# 6 clusters for DIANA
ggthemr("flat", spacing = 0.1)

di_labcolors <- unsup_labs_reord_di %>% 
  as_tibble() %>% 
  mutate(labcolor = case_when(
    BMI_cat == "UW" ~ "#355E44",
    BMI_cat == "HW" ~ "#34495E",
    BMI_cat == "OW" ~ "#3F345E",
    BMI_cat == "OB" ~ "#5E343F",
  ))

diana_dend_clust_init <- nhanes_diana %>% 
  as.dendrogram() %>% 
  set("leaves_pch", 19, col = di_labcolors$leafcolor) %>% 
  set("leaves_col", di_labcolors$leafcolor) %>% 
  set("branches_k_col", k = 6, c("#3498DB",    # Blue
                                 "#2ECC71",    # Green
                                 "#F39C12",    # Gold/orange
                                 "#E74C3C",    # Red
                                 "#D979C9",    # Pink
                                 "#9B59B6"))#,    # Purple
                                 #"#9F83D4",    # Grape
                                 #"#1ABCB4",    # Teal
                                 #"#34495E"))   # Slate blue

diana_dend_clust_init_ggd <- diana_dend_clust_init %>% 
  as.ggdend()

diana_dend_clust_init_gg <- ggplot(diana_dend_clust_init_ggd, labels = FALSE) +
  scale_y_reverse() +
  coord_polar(theta = "x") +
  theme(axis.line = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  xlab("") +
  ylab("") #+
  #ggtitle("DIANA Dendrogram, k = 6 Clusters")
diana_dend_clust_init_gg

ggthemr("flat")
```

Though there are some clusters in each of these dendrograms having low heterogeneity among BMI category labels, the plots indicate that there are an insufficient number of clusters to discriminate well among the groups.  The next step is to rerun each of these dendrograms using higher cluster numbers.








<!-- CONTINUE HERE: -->
<!-- The below text needs to be updated after re-evaluation of these initial dendrogram plots. -->
<!-- At first glance it seems both plots were able to group obese and overweight pretty well for one cluster! -->












<!-- These dendrograms were re-run using $k = 6$ clusters for AGNES and $k = 9$ clusters for DIANA; output is shown below. -->

<!-- ```{r agnes_dend_clust_v2, echo = FALSE, warning = FALSE, paged.print = FALSE, fig.width = 11, fig.height = 11} -->
<!-- # 6 clusters for AGNES -->
<!-- ggthemr("flat", spacing = 0.1) -->

<!-- agnes_dend_clust_v2 <- nhanes_agnes %>%  -->
<!--   as.dendrogram() %>%  -->
<!--   set("leaves_pch", 19, col = ag_labcolors$leafcolor) %>%  -->
<!--   set("leaves_col", ag_labcolors$leafcolor) %>%  -->
<!--   set("branches_k_col", k = 6, c("#3498DB",    # Blue -->
<!--                                  "#2ECC71",    # Green -->
<!--                                  "#F39C12",    # Gold/orange -->
<!--                                  "#E74C3C",    # Red -->
<!--                                  #"#D979C9",    # Pink -->
<!--                                  "#9B59B6",    # Purple -->
<!--                                  #"#9F83D4",    # Grape -->
<!--                                  #"#1ABCB4",    # Teal -->
<!--                                  "#34495E"))   # Slate blue -->

<!-- agnes_dend_clust_v2_ggd <- agnes_dend_clust_v2 %>%  -->
<!--   as.ggdend() -->

<!-- agnes_dend_clust_v2_gg <- ggplot(agnes_dend_clust_v2_ggd, labels = FALSE) + -->
<!--   scale_y_reverse() + -->
<!--   coord_polar(theta = "x") + -->
<!--   theme(axis.line = element_blank(), -->
<!--         axis.text.x = element_blank(), -->
<!--         axis.text.y = element_blank(), -->
<!--         axis.ticks = element_blank(), -->
<!--         panel.grid = element_blank()) + -->
<!--   xlab("") + -->
<!--   ylab("") #+ -->
<!--   #ggtitle("AGNES Dendrogram, k = 6 Clusters") -->
<!-- agnes_dend_clust_v2_gg -->

<!-- ggthemr("flat") -->
<!-- ``` -->

<!-- ```{r diana_dend_clust_v2, echo = FALSE, warning = FALSE, paged.print = FALSE, fig.width = 11, fig.height = 11} -->
<!-- # 9 clusters for DIANA -->
<!-- ggthemr("flat", spacing = 0.1) -->

<!-- diana_dend_clust_v2 <- nhanes_diana %>%  -->
<!--   as.dendrogram() %>%  -->
<!--   set("leaves_pch", 19, col = di_labcolors$leafcolor) %>%  -->
<!--   set("leaves_col", di_labcolors$leafcolor) %>%  -->
<!--   set("branches_k_col", k = 9, c("#3498DB",    # Blue -->
<!--                                  "#2ECC71",    # Green -->
<!--                                  "#F39C12",    # Gold/orange -->
<!--                                  "#E74C3C",    # Red -->
<!--                                  "#D979C9",    # Pink -->
<!--                                  "#9B59B6",    # Purple -->
<!--                                  "#9F83D4",    # Grape -->
<!--                                  "#1ABCB4",    # Teal -->
<!--                                  "#34495E"))   # Slate blue -->

<!-- diana_dend_clust_v2_ggd <- diana_dend_clust_v2 %>%  -->
<!--   as.ggdend() -->

<!-- diana_dend_clust_v2_gg <- ggplot(diana_dend_clust_v2_ggd, labels = FALSE) + -->
<!--   scale_y_reverse() + -->
<!--   coord_polar(theta = "x") + -->
<!--   theme(axis.line = element_blank(), -->
<!--         axis.text.x = element_blank(), -->
<!--         axis.text.y = element_blank(), -->
<!--         axis.ticks = element_blank(), -->
<!--         panel.grid = element_blank()) + -->
<!--   xlab("") + -->
<!--   ylab("") #+ -->
<!--   #ggtitle("DIANA Dendrogram, k = 9 Clusters") -->
<!-- diana_dend_clust_v2_gg -->

<!-- ggthemr("flat") -->
<!-- ``` -->